A Hedonic Price Analysis of Traffic Noise in the Twin Cities Housing Market

### 2013-09-15 reports/Draft.Rnw
added global regression table
few more edits to intro

\newcolumntype{x}[1]{%
>{\centering\hspace{0pt}}p{#1}}%

@@

\doublespacing
\begin{document}
\maketitle
\begin{singlespace}
\begin{abstract}
One consequence of the expanding road network and its associated traffic is increased levels of traffic noise.  While the hedonic literature has consistently shown a negative effect of this phenomenon on the real estate market, the relationship between housing prices and automobile traffic noise has rarely been explored at landscape level, especially in the United States. Here, we show this relationship by modeling the propagation of traffic noise over the landscape and by analyzing the extent of its impact on single-family home transactions in the urban area surrounding St. Paul, Minnesota using spatially explicit local regression techniques. Our estimate of the negative relationship between traffic noise levels and house prices is smaller in magnitude than most of the relevant published literature. We find no evidence that the effect of traffic noise varies significantly by geography, nor by traffic noise levels.  We find significant differences in the impact of traffic noise before and after the economic recession of 2008-09. Our results suggest that an increase in traffic noise of one decibel decreases house prices by an average of 0.19 percent before September 2008, and an average of 0.37 percent after September 2008.
\end{abstract}
\end{singlespace}

@@

Reviews of the research by \citet{Bateman2001} found that the negative effect of traffic noise on house prices ranged from 0.08 to 2.22 percent per decibel with a mean around 0.55 percent per decibel, while \citet{Navrud2002} found an average decrease of 0.64 percent per decibel. However, much of the research in these reviews and subsequent work has tended to focus on the European experience with noise, especially the United Kingdom \citep{Day2007, Blanco2011}, Sweden \citep{Wilhelmsson2000, Andersson2010}, Switzerland \citep{Baranzini2010}, Spain \citep{MarmolejoDuarteCarlos;GonzalezTamez2009}, Netherlands \citep{Theebe2004a}, and Germany \citep{Brandt2011}.

@@

In this paper, we seek to contribute to the noise hedonic literature by examining the relationship between single family home prices from 2005 to 2010 and variation in road traffic noise exposure in the St.\ Paul, Minnesota urban area. Our work makes multiple contributions to the field. First, we utilize new estimates of traffic noise that has never before been used in the hedonic literature. The noise propagation model estimates traffic noise levels at a high spatial resolution (10m x 10m) while also incorporating the effects of nearby buildings and landcover.

@@

Lastly, we present the first estimates of the implicit price of traffic noise after the Great Recession and the housing bubble collapse of 2008-09. Contrary to work such as \citet{Cho2011b}, which found that environmental amenities mattered less after the housing crash, our estimated effect of traffic noise mattered more after the recession (0.37 percent reduction in house prices per additional decibel) compared to beforehand (0.19 percent per decibel).

@@

\makebox[\textwidth][c]{\includegraphics[width = 1.3\textwidth]{../graphs/GlobalRegression}}
\end{table}

Standard urban economic theory \citep{Alonso1964, Mills1967, Muth1969} predicts that the price of land will vary over space within our dataset to account for locational amenities. As such, we add interaction terms between lot size and distance to the nearest central business district to account for spatial variation in the price of land. The first column in Table \ref{tab:globalRegression} shows the results of this regression. The negative coefficient of -0.0027 on the noise variable (a one dB increase in noise decreases house price by 0.27 percent) is in line with the estimates described earlier in the literature. The significant coefficients on the lot size interaction terms suggest that the value of land varies over space.

Table \ref{tab:globalRegression} also attempts to begin to understand how the economic recession may have influenced the hedonic function in our study area. The other columns in the table show the regression output from separating the data into sales before and after September of 2008 (the month that the US Federal government took over Fannie Mae and Freddie Mac, Lehman Brothers filed for bankruptcy, and the American International Group (AIG) narrowly avoided bankruptcy only through a \$85 billion loan from the US Federal government).\footnote{\url{http://www.federalreserve.gov/newsevents/press/other/20080916a.htm}} The coefficients for some of the hedonic variables of interest (noise, schools, land, for instance) appear different across the two time periods.

>>>>

\newcommand\T{\rule{0pt}{2.6ex}}       % Top strut
\newcommand\B{\rule[-1.2ex]{0pt}{0pt}} % Bottom strut
\newcolumntype{x}[1]{%
>{\centering\hspace{0pt}}p{#1}}%

@@

\doublespacing
\begin{document}
\maketitle
\begin{singlespace}
\begin{abstract}
One consequence of the expanding road network and its associated traffic is increased levels of traffic noise.  While the hedonic literature has consistently shown a negative effect of this phenomenon on the real estate market, the relationship between housing prices and automobile traffic noise has rarely been explored at landscape level, especially in the United States. Here, we show this relationship by modeling the propagation of traffic noise over the landscape and by analyzing the extent of its impact on single-family home transactions in the urban area surrounding St.\ Paul, Minnesota using spatially explicit local regression techniques. Our estimate of the negative relationship between traffic noise levels and house prices is smaller in magnitude than most of the relevant published literature. We find no evidence that the effect of traffic noise varies significantly by geography, nor by traffic noise levels.  We find significant differences in the impact of traffic noise before and after the economic recession of 2008-09. Our results suggest that an increase in traffic noise of one decibel decreases house prices by an average of 0.19 percent before September 2008, and an average of 0.37 percent after September 2008.
\end{abstract}
\end{singlespace}

@@

Reviews of the research by \citet{Bateman2001} found that the negative effect of traffic noise on house prices ranged from 0.08 to 2.22 percent per decibel with a mean around 0.55 percent per decibel, while \citet{Navrud2002} found an average decrease of 0.64 percent per decibel. Much of the research in this literature has tended to focus on the European experience with noise, especially the United Kingdom \citep{Day2007, Blanco2011}, Sweden \citep{Wilhelmsson2000, Andersson2010}, Switzerland \citep{Baranzini2010}, Spain \citep{MarmolejoDuarteCarlos;GonzalezTamez2009}, Netherlands \citep{Theebe2004a}, and Germany \citep{Brandt2011}.

@@

In this paper, we contribute to the noise hedonic literature by examining the relationship between single family home prices from 2005 to 2010 and variation in road traffic noise exposure in the St.\ Paul, Minnesota urban area across 44 different cities. First, we utilize new estimates of traffic noise that have never before been used in the hedonic literature. The noise propagation model estimates traffic noise levels at a high spatial resolution (10m x 10m) while also incorporating the effects of nearby buildings and landcover.

@@

Lastly, we present the first estimates of the implicit price of traffic noise after the Great Recession and the housing bubble collapse of 2008-09. Contrary to work such as \citet{Cho2011b}, which found that environmental amenities mattered less after the housing crash, our estimated effect of traffic noise mattered more after the recession (0.37 percent reduction in house prices per additional decibel) compared to only 0.19 percent per decibel beforehand.

@@

\makebox[\textwidth][c]{%\includegraphics[width = 1.3\textwidth]{../graphs/GlobalRegression}}
% Table generated by Excel2LaTeX from sheet 'Sheet1'
  \centering
  \tabcolsep=0.03cm
    \begin{tabular}{lclcclcclccl}
   & (1)   &       &       & (2)   &       &       & (3)   &       &       & (4)   &  \\ \hline
\T  Intercept & 9.37E+00 & ***   & \hspace{.5cm} & 9.49E+00 & ***   & \hspace{.5cm}   & 8.55E+00 & ***   & \hspace{.5cm}   & 1.10E+01 & *** \\
          & (9.48E-02) &    &   & (9.62E-02) &   &   & (1.11E-01) &       &       & (2.12E-01) &  \\
\T   House Size & 3.38E-04 & ***   &   & 3.39E-04 & ***   &   & 3.35E-04 & ***   &   & 3.50E-04 & *** \\
          & (1.85E-06) &   &   & (1.86E-06) &       &       & (1.99E-06) &       &       & (4.26E-06) &  \\
\T  Lot Size & 1.42E-01 & ***   &    & 2.74E-01 & ***   &   & 2.31E-01 & ***   &      & 2.74E-01 & *** \\
          & (1.14E-02) &     &   & (2.12E-02) &     &      & (2.29E-02) &       &       & (4.99E-02) &  \\
\T  Owner Occupancy & 3.22E-02 & ***   &   & 3.23E-02 & ***   &     & 3.38E-02 & ***  & & 2.61E-02 & *** \\
          & (2.80E-03) &  &   & (2.79E-03) &       &       & (3.24E-03) &       &       & (6.41E-03) &  \\
\T  Year Built & 7.82E-04 & ***   &   & 7.17E-04 & ***   &    & 9.68E-04 & ***   &   & 2.15E-06 &  \\
          & (4.57E-05) &       &       & (4.65E-05) &   &   & (5.06E-05) &    &   & (1.05E-04) &  \\
\T  Traffic Noise & -2.02E-03 & *** &   & -2.00E-03 & ***   &   & -1.63E-03 & ***  & & -3.08E-03 & *** \\
          & (1.32E-04) &   &   & (1.32E-04) &       &       & (1.38E-04) &       &       & (3.29E-04) &  \\
\T  Median Income & 3.03E-06 & ***  & & 3.31E-06 & ***  &   & 2.86E-06 & ***   &       & 4.06E-06 & *** \\
          & (5.87E-08) &       &       & (5.87E-08) &   &   & (6.56E-08) &       &       & (1.35E-07) &  \\
\T  Elem. Test Scores & 1.88E-03 & *** &   & 1.81E-03 & ***   &   & 3.11E-03 & *** & & 1.21E-03 & *** \\
          & (1.06E-04) &  &   & (1.06E-04) &   &       & (1.75E-04) &       &       & (1.55E-04) &  \\
\T  Distance to CBD & -1.56E-06 & ***  &  & 7.46E-07 & *     &  & -1.57E-07 &   &       & 1.40E-06 &  \\
          & (1.97E-07) &       &       & (3.70E-07) &       &       & (3.95E-07) &   &   & (8.98E-07) &  \\
\T  Distance to Park & -2.38E-07 &   &    & -5.48E-07 &    &   & -6.37E-07 &  &   & -4.01E-07 &  \\
          & (6.82E-07) & &   & (6.83E-07) &       &       & (7.33E-07) &       &       & (1.58E-06) &  \\
\T  Distance to Lake & 2.12E-05 & ***   &   & 2.36E-05 & ***   &   & 2.22E-05 & ***   & & 2.79E-05 & *** \\
          & (1.68E-06) &       &    & (1.71E-06) &   &    & (1.83E-06) & &  & (4.03E-06) &  \\
\T  Distance to Shop & 3.01E-06 & ***   &    & 3.12E-06 & ***   & & 4.16E-06 & ***   &    & 1.88E-06 &  \\
          & (7.67E-07) &       &       & (7.66E-07) &       &       & (8.17E-07) &  & & (1.82E-06) &  \\
\T  Lot size * Distance to CBD &  &   & & -8.81E-06 & *** & & -6.66E-06 & ***   &       & -9.55E-06 & ** \\
          &   &   &   & (1.19E-06) &       &       & (1.27E-06) &       &       & (2.94E-06) &  \\
\T  County Fixed Effects & yes   &   &   & yes   &       &       & yes   &       &       & yes   &  \\
    Year of Sale Fixed Effects & yes   &   &   & yes   &   &       & yes   &       &       & yes   &  \\
    Month of Sale Fixed Effects & yes   &   &   & yes   &   &       & yes   &       &       & yes   &  \\ \hline
\T  Number of Observations &        42,095  &   &  &    42,095  & & &   31,253  & & &    10,842  &  \\
    R$^2$   & 0.696 &       &       & 0.696 &       &       & 0.714 &       &       & 0.613 &  \\
 \multicolumn{12}{c}{(standard errors) ***$p<.001$; **$p<.01$; *$p<.05$}  \\
    \end{tabular}%
}
\end{table}

The results of four regression models are displayed in Table \ref{tab:globalRegression}. The first column implements a basic regression as described in equation \eqref{eq:model} and estimates the impact of traffic noise to be roughly 0.002, or a 0.2 percent reduction in house prices for each additional decibel of traffic noise. Standard urban economic theory \citep{Alonso1964, Mills1967, Muth1969} predicts that the price of land will vary over space within our dataset to account for locational amenities. As such, we add interaction terms between lot size and distance to the Central Business District to account for spatial variation in the price of land in the second column in Table \ref{tab:globalRegression}. The significant coefficients on the interaction term suggest that the value of land varies over space. We also find significant coefficients in regressions including interaction terms between quadratic forumations of distance to the Central Business District and lot size, thereby suggesting that the relationship between lot size and house prices exhibits non-linearities over space.

Table \ref{tab:globalRegression} also attempts to begin to understand how the economic recession may have influenced the hedonic function in our study area. Columns (3) and (4) in the table show the regression output from separating the data into sales before and after September of 2008 (the month that the US Federal government took over Fannie Mae and Freddie Mac, Lehman Brothers filed for bankruptcy, and the American International Group (AIG) narrowly avoided bankruptcy only through a \$85 billion loan from the US Federal government).\footnote{\url{http://www.federalreserve.gov/newsevents/press/other/20080916a.htm}} The coefficients for some of the hedonic variables of interest (noise, income, and schools, for instance) appear different across the two time periods.
